---
title: "Multi-sample comparisons of the Hayman/Pasricha (C084) ZIPT mini-bulk data set"
description: |
author:
  - name: Peter Hickey
    url: https://peterhickey.org
    affiliation: WEHI SCORE
    affiliation_url: https://www.wehi.edu.au/people/shalin-naik/3310/score
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
bibliography: ref.bib
---

```{r setup}
library(SingleCellExperiment)
library(here)
library(edgeR)
library(BiocParallel)
library(scales)
library(ggplot2)
library(cowplot)
library(patchwork)
library(pheatmap)

# NOTE: Using >= 4 cores seizes up my laptop. Can use more on RStudio server.
options(
  "mc.cores" = 
    ifelse(Sys.info()[["nodename"]] == "rstudio-1.hpc.wehi.edu.au", 8L, 2L))

register(MulticoreParam(workers = getOption("mc.cores")))

source(here("code", "helper_functions.R"))

knitr::opts_chunk$set(
  fig.path = "C084_Hayman_Pasricha.multi-sample_comparisons_files/")
```

# Preparing the data

We start from the preprocessed *SingleCellExperiment* object created in ['Preprocessing the Hayman/Pasricha (C084) ZIPT mini-bulk data sett'](C084_Hayman_Pasricha.preprocess.html).

```{r}
sce <- readRDS(
  here("data", "SCEs", "C084_Hayman_Pasricha.preprocessed.SCE.rds"))

# Some useful colours
plate_number_colours <- setNames(
  unique(sce$plate_number_colours),
  unique(names(sce$plate_number_colours)))
treatment_colours <- setNames(
  unique(sce$treatment_colours),
  unique(names(sce$treatment_colours)))
timepoint_colours <- setNames(
  unique(sce$timepoint_colours),
  unique(names(sce$timepoint_colours)))
sex_colours <- setNames(
  unique(sce$sex_colours),
  unique(names(sce$sex_colours)))

# Some useful gene sets
mito_set <- rownames(sce)[any(rowData(sce)$ENSEMBL.SEQNAME == "MT")]

ribo_set <- grep("^RP(S|L)", rownames(sce), value = TRUE)
# NOTE: A more curated approach for identifying ribosomal protein genes 
#       (https://github.com/Bioconductor/OrchestratingSingleCellAnalysis-base/blob/ae201bf26e3e4fa82d9165d8abf4f4dc4b8e5a68/feature-selection.Rmd#L376-L380)
library(msigdbr)
c2_sets <- msigdbr(species = "Homo sapiens", category = "C2")
ribo_set <- union(
  ribo_set,
  c2_sets[c2_sets$gs_name == "KEGG_RIBOSOME", ]$gene_symbol)

e <- new.env()
load(url("http://bioinf.wehi.edu.au/software/GenderGenes/GenderGenes.RData"), e)
sex_set <- rownames(sce)[
  any(rowData(sce)$NCBI.ENTREZID %in% c(e$XiEgenes, e$msYgenes))]

pseudogene_set <- rownames(sce)[
  any(grepl("pseudogene", rowData(sce)$ENSEMBL.GENEBIOTYPE))]

hemo_set <- rownames(sce)[
  any(grepl("hemoglobin subunit", rowData(sce)$NCBI.GENENAME))]
```

We convert the data to a *DGEList* object in preparation for analysis with `r BiocStyle::Biocpkg("edgeR")`.

```{r}
x <- DGEList(
  counts = as.matrix(counts(sce)),
  samples = colData(sce),
  group = factor(paste0(sce$treatment, ".", sce$timepoint)),
  genes = flattenDF(rowData(sce)))
x_lc476 <- x[, x$samples$plate_number == "LC476"]
```

# Data pre-processing

## Transformations from the raw-scale

For differential expression and related analyses, gene expression is rarely considered at the level of raw counts since libraries sequenced at a greater depth will result in higher counts.
Rather, it is common practice to transform raw counts onto a scale that accounts for such library size differences.
Popular transformations include counts per million (CPM), log2-counts per million (log-CPM), reads per kilobase of transcript per million (RPKM), and fragments per kilobase of transcript per million (FPKM).

In our analyses, CPM and log-CPM transformations are used regularly although they do not account for gene length differences as RPKM and FPKM values do. 
Whilst RPKM and FPKM values can just as well be used, CPM and log-CPM values can be calculated using a counts matrix alone and will suffice for the type of comparisons we are interested in^[Furthermore, the 3'-tag sequencing of SCORE's mini-bulk protocol basically negates the need to account for gene length differences.].
Assuming that there are no differences in isoform usage between conditions, differential expression analyses look at gene expression changes between conditions rather than comparing expression across multiple genes or drawing conclusions on absolute levels of expression.
In other words, gene lengths remain constant for comparisons of interest and any observed differences are a result of changes in condition rather than changes in gene length.

```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log = TRUE)
```

A CPM value of 1 for a gene equates to having `r number(min(x$samples$lib.size / 1e6), accuracy = 0.01)` counts in the sample with the lowest sequencing depth (`r colnames(x)[which.min(x$samples$sum)]`, library size approx. `r number(min(x$samples$lib.size / 1e6), accuracy = 0.01)` million) or `r number(max(x$samples$lib.size / 1e6), accuracy = 0.01)` counts in the sample with the greatest sequencing depth (`r colnames(x)[which.max(x$samples$sum)]`, library size approx. `r number(max(x$samples$lib.size / 1e6), accuracy = 0.01)` million).

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6

L_lc476 <- mean(x_lc476$samples$lib.size) * 1e-6
M_lc476 <- median(x_lc476$samples$lib.size) * 1e-6
```

The log-CPM values will be used for exploratory plots.
The log-CPM values use an offset to the CPM values before converting to the log2-scale.
By default, the offset is $2/L$ where 2 is the 'prior count' and $L$ is the average library size in millions, so the log-CPM values are related to the CPM values by $log_{2}(CPM + 2/L)$.
This calculation ensures that any two read counts with identical CPM values will also have identical log-CPM values.
The prior count avoids taking the logarithm of zero, and also reduces spurious variability for genes with very low counts by shrinking all the inter-sample log-fold-changes towards zero, something that is helpful for exploratory plotting.

For this dataset, the average library size is about `r number(L, accuracy = 0.01)` million for the entire dataset (`r number(L_lc476, accuracy = 0.01)` million for plate `LC476`), so $L \approx$ `r number(L, accuracy = 0.01)` (resp. `r number(L_lc476, accuracy = 0.01)`) and the minimum log-CPM value for each sample becomes $log_{2}(CPM + 2/L) \approx$ `r number(log2(2 / L), accuracy = 0.01)` (resp. `r number(log2(2 / L_lc476), accuracy = 0.01)`).
In other words, a count of zero for this data maps to a log-CPM value of `r number(log2(2 / L), accuracy = 0.01)` (resp. `r number(log2(2 / L_lc476), accuracy = 0.01)`) after adding the prior count or offset.

<aside>
The values in parenthesis refer to those for just [Using LC476].
</aside>

## Removing genes that are lowly expressed

```{r}
stopifnot(sum(matrixStats::rowAlls(x$counts == 0)) == 0)
```

All datasets will include a mix of genes that are expressed and those that are not expressed.
Whilst it is of interest to examine genes that are expressed in one condition but not in another, some genes are unexpressed throughout all samples.
We have already removed all genes in this dataset that have zero counts across all `r ncol(x)` samples.

Figure \@ref(fig:lcpm-density) plots the distribution log-CPM values and shows that a sizeable proportion of genes within each sample are either unexpressed or lowly-expressed with log-CPM values that are small.
Genes that do not have a worthwhile number of reads in any sample should be filtered out of the downstream analyses.
There are several reasons for this.
From a biological point of view, genes that not expressed at a biologically meaningful level in any condition are not of interest and are therefore best ignored.
From a statistical point of view, removing low count genes allows the mean-variance relationship in the data to be estimated with greater reliability and also reduces the number of statistical tests that need to be carried out in downstream analyses looking at differential expression.

The `filterByExpr()` function in the `r BiocStyle::Biocpkg("edgeR")` package provides an automatic way to filter genes, while keeping as many genes as possible with worthwhile counts.

```{r}
lcpm <- cpm(x, log = TRUE)
keep_exprs <- filterByExpr(x, group = x$samples$group)
x <- x[keep_exprs, , keep.lib.sizes = FALSE]

lcpm_lc476 <- cpm(x_lc476, log = TRUE)
keep_exprs_lc476 <- filterByExpr(x_lc476, group = x_lc476$samples$group)
x_lc476 <- x_lc476[keep_exprs_lc476, , keep.lib.sizes = FALSE]
```

By default, `filterByExpr()` keeps genes with about 10 read counts or more in a minimum number of samples, where the number of samples is chosen according to the minimum group sample size^[The actual filtering uses CPM values rather than counts in order to avoid giving preference to samples with large library sizes.].
For this dataset, the median library size is about `r number(M, accuracy = 0.01)` million (resp. `r number(M_lc476, accuracy = 0.01)` million) and 10 / `r number(M, accuracy = 0.01)` approx. `r number(10 / M, accuracy = 0.1)` (resp. 10 / `r number(M_lc476, accuracy = 0.01)` approx. `r number(10 / M_lc476, accuracy = 0.1)`), so the `filterByExpr()` function keeps genes that have a CPM of `r number(10 / M, accuracy = 0.1)` (resp. `r number(10 / M_lc476, accuracy = 0.1)`) or more in at least `r min(table(x$samples$group))` (resp. `r min(table(x_lc476$samples$group))`) samples.
A biologically interesting gene should be expressed in at least `r min(table(x$samples$group))` (resp. `r min(table(x_lc476$samples$group))`) samples because all the groups have at least `r min(table(x$samples$group))` (resp. `r min(table(x_lc476$samples$group))`) replicates within each plate.
The cutoffs used depend on the sequencing depth and on the experimental design. If the library sizes had been larger then a lower CPM cutoff would have been chosen, because larger library sizes provide better resolution to explore more genes at lower expression levels.
Alternatively, smaller library sizes decrease our ability to explore marginal genes and hence would have led to a higher CPM cutoff.

<aside>
The values in parenthesis refer to those for just [Using LC476].
</aside>

Using this criterion, the number of genes is reduced to `r number(sum(keep_exprs), big.mark = ",")` (resp. `r number(sum(keep_exprs_lc476), big.mark = ",")`), about `r percent(sum(keep_exprs) / length(keep_exprs))` (resp. `r percent(sum(keep_exprs_lc476) / length(keep_exprs_lc476))`) of the number that we started with (right panels of Figure \@ref(fig:lcpm-density)).

<aside>
The values in parenthesis refer to those for just [Using LC476].
</aside>

```{r lcpm-density, fig.cap = "The density of log-CPM values for raw pre-filtered data (left) and post-filtered data (right) are shown for 12 randomly chosen samples. Dotted vertical lines mark the log-CPM threshold used in the filtering step.", fig.asp = 1}
f <- function(lcpm, x, j, samplenames, col, lcpm_cutoff, prefix) {
  plot(
    density(lcpm[, j[1]]),
    col = col[1],
    lwd = 2,
    ylim = c(0, 2.6),
    las = 2,
    main = "",
    xlab = "",
    xlim = c(0, 14))
  title(main = paste0(prefix, ": Raw data"), xlab = "Log-cpm")
  abline(v = lcpm_cutoff, lty = 3)
  for (i in seq_along(j)[-1]) {
    den <- density(lcpm[, j[i]])
    lines(den$x, den$y, col = col[i], lwd = 2)
  }
  legend("topright", samplenames, text.col = col, bty = "n", cex = 0.5, ncol = 2)
  
  lcpm <- cpm(x, log = TRUE)
  plot(
    density(lcpm[, 1]),
    col = col[1],
    lwd = 2,
    ylim = c(0, 2.6),
    las = 2,
    main = "",
    xlab = "",
    xlim = c(0, 14))
  title(main = paste0(prefix, ": Filtered data"), xlab = "Log-cpm")
  abline(v = lcpm_cutoff, lty = 3)
  for (i in seq_along(j)[-1]) {
    den <- density(lcpm[, j[i]])
    lines(den$x, den$y, col = col[i], lwd = 2)
  }
  legend("topright", samplenames, text.col = col, bty = "n", cex = 0.5, ncol = 2)
}

set.seed(666)
nsamples <- 12
col <- palette.colors(nsamples, "Polychrome 36")
j <- sort(sample(ncol(x), nsamples))
samplenames <- colnames(x)[j]
j_lc476 <- sort(sample(ncol(x_lc476), nsamples))
samplenames_lc476 <- colnames(x)[j_lc476]
lcpm_cutoff <- log2(10 / M + 2 / L)
lcpm_cutoff_lc476 <- log2(10 / M_lc476 + 2 / L_lc476)

par(mfrow = c(2, 2))
f(lcpm, x, j, samplenames, col, lcpm_cutoff, "Both plates")
f(lcpm_lc476, x_lc476, j_lc476, samplenames_lc476, col, lcpm_cutoff_lc476, "LC476")
```

## Normalising gene expression distributions

During the sample preparation or sequencing process, external factors that are not of biological interest can affect the expression of individual samples.
For example, samples processed in the first batch of an experiment can have higher expression overall when compared to samples processed in a second batch. 
It is assumed that all samples should have a similar range and distribution of expression values.
Normalisation is required to ensure that the expression distributions of each sample are similar across the entire experiment.

Any plot showing the per sample expression distributions, such as a density or boxplot, is useful in determining whether any samples are dissimilar to others. 
Distributions of log-CPM values appear to be similar throughout all samples within this dataset (right panels of Figure \@ref(fig:lcpm-density)).

Nonetheless, normalisation by the method of trimmed mean of M-values (TMM) [@robinson2010scaling] is performed using the `calcNormFactors()` function in `r BiocStyle::Biocpkg("edgeR")`.
The normalisation factors calculated here are used as a scaling factor for the library sizes.

```{r}
x <- calcNormFactors(x, method = "TMM")
x_lc476 <- calcNormFactors(x_lc476, method = "TMM")
```

For this dataset the effect of TMM-normalisation is mild, as evident in the magnitude of the scaling factors, which are all relatively close to 1 in both the overall dataset (median = `r number(median(x$samples$norm.factors), 0.01)`, IQR = `r number(iqr(x$samples$norm.factors), 0.01)`, 5% and 95% quantiles = `r glue::glue_collapse(number(quantile(x$samples$norm.factors, c(0.05, 0.95)), accuracy = 0.01), sep = " and ")`) and when restricted to just plate `LC476` (median = `r number(median(x_lc476$samples$norm.factors), 0.01)`, IQR = `r number(iqr(x_lc476$samples$norm.factors), 0.01)`, 5% and 95% quantiles = `r glue::glue_collapse(number(quantile(x_lc476$samples$norm.factors, c(0.05, 0.95)), accuracy = 0.01), sep = " and ")`)

Figure \@ref(fig:lcpm-boxplot) shows the expression distribution of 12 randomly chosen samples for unnormalised and normalised data, where distributions are noticeably different pre-normalisation and are similar post-normalisation.

```{r lcpm-boxplot, fig.cap = "Boxplots of log-CPM values showing expression distributions for unnormalised data (left) and normalised data (right) for 12 randomly chosen samples.", fig.asp = 1}
g <- function(lcpm, x, j, samplenames, col, prefix) {
  boxplot(lcpm[, j], las = 2, col = col, main = "", cex = 0.1, xaxt = "n")
  title(main = paste0(prefix, ": Unnormalised data"), ylab = "Log-cpm")
  
  lcpm <- cpm(x, log = TRUE)
  boxplot(lcpm[, j], las = 2, col = col, main = "", cex = 0.1, xaxt = "n")
  title(main = paste0(prefix, ": Normalised data"), ylab = "Log-cpm")
}

col <- palette.colors(nsamples, "Polychrome 36")
nsamples <- 12
set.seed(666)
j <- sort(sample(ncol(x), nsamples))
samplenames <- colnames(x)[j]
j_lc476 <- sort(sample(ncol(x_lc476), nsamples))
samplenames_lc476 <- colnames(x_lc476)[j_lc476]

par(mfrow = c(2, 2))
g(lcpm, x, j, samplenames, col, "Both plates")
g(lcpm_lc476, x_lc476, j_lc476, samplenames_lc476, col, "LC476")
```

## Unsupervised clustering of samples

In our opinion, one of the most important exploratory plots to examine for gene expression analyses is the multi-dimensional scaling (MDS) plot, or similar.
The plot shows similarities and dissimilarities between samples in an unsupervised manner so that one can have an idea of the extent to which differential expression can be detected before carrying out formal tests.
Ideally, samples would cluster well within the primary condition of interest, and any sample straying far from its group could be identified and followed up for sources of error or extra variation.
If present, technical replicates should lie very close to one another.

Such a plot can be made in `r BiocStyle::Biocpkg("limma")` using the `plotMDS()` function.
The first dimension represents the leading-fold-change that best separates samples and explains the largest proportion of variation in the data, with subsequent dimensions having a smaller effect and being orthogonal to the ones before it. 
When experimental design involves multiple factors, it is recommended that each factor is examined over several dimensions.
If samples cluster by a given factor in any of these dimensions, it suggests that the factor contributes to expression differences and is worth including in the linear modelling.
On the other hand, factors that show little or no effect may be left out of downstream analysis.

Figure \@ref(fig:mds) shows that samples can be seen to cluster strongly by plate number over dimension 1.
This dominates any variation due to timepoint or treatment.

```{r mds, fig.cap = "MDS plots of log-CPM values over dimensions 1 and 2 and dimensioned 3 and 4 with samples coloured by batch, treatment, and timepoint. Distances on the plot correspond to the leading fold-change, which is the average (root-mean-square) log2-fold-change for the 500 genes most divergent between each pair of samples by default.", fig.asp = 3 / 2}
h <- function(lcpm, col_list, location) {
  for (i in seq_along(col_list)) {
    plotMDS(lcpm, pch = 16, col = col_list[[i]])
    legend(
      location,
      col = unique(col_list[[i]]),
      legend = unique(names(col_list[[i]])),
      pch = 16,
      title = names(col_list)[i])
    plotMDS(lcpm, pch = 16, col = col_list[[i]], dim = c(3, 4))
  }
}

lcpm <- cpm(x, log = TRUE)
col_list <- list(
  "plate_number" = plate_number_colours[x$samples$plate_number],
  "treatment" = treatment_colours[x$samples$treatment],
  "timepoint" = timepoint_colours[x$samples$timepoint])
par(mfrow = c(3, 2))
h(lcpm, col_list, "bottom")
```

The `removeBatchEffect()` function in the `r BiocStyle::Biocpkg("limma")` package provides a way to remove the 'plate number effect' (i.e. the batch effect) for the purposes of visualization.
Figure \@ref(fig:mds-removeBatchEffect) shows that, once the batch effect is accounted for, we observe some clustering along dimension 1 by timepoint but little association with treatment over dimensions $1-4$.

```{r mds-removeBatchEffect, fig.cap = "MDS plots of log-CPM values after removing the batch effect over dimensions 1 and 2 and dimensioned 3 and 4 with samples coloured by batch, treatment, and timepoint. Distances on the plot correspond to the leading fold-change, which is the average (root-mean-square) log2-fold-change for the 500 genes most divergent between each pair of samples by default.", fig.asp = 3 / 2}
lcpm_rbe <- removeBatchEffect(
  lcpm,
  batch = x$samples$plate_number,
  design = model.matrix(~0 + group, x$samples))
col_list <- list(
  "plate_number" = plate_number_colours[x$samples$plate_number],
  "treatment" = treatment_colours[x$samples$treatment],
  "timepoint" = timepoint_colours[x$samples$timepoint])
par(mfrow = c(3, 2))
h(lcpm_rbe, col_list, "topleft")
```

Figures \@ref(fig:mds-lc475) and \@ref(fig:mds-lc476) show MDS plots of the uncorrected data when restricted to plate `LC475` or `LC476`, respectively.
In each case, we observe some clustering along dimension 1 by timepoint but little association with treatment over dimensions $1-4$.

```{r mds-lc475, fig.cap = "MDS plots of `LC475` log-CPM values over dimensions 1 and 2 and dimensioned 3 and 4 with samples coloured by treatment and timepoint. Distances on the plot correspond to the leading fold-change, which is the average (root-mean-square) log2-fold-change for the 500 genes most divergent between each pair of samples by default.", fig.asp = 1}
x_lc475 <- x[, x$samples$plate_number == "LC475"]
lcpm_lc475 <- cpm(x_lc475, log = TRUE)

col_list <- list(
  "treatment" = treatment_colours[x_lc475$samples$treatment],
  "timepoint" = timepoint_colours[x_lc475$samples$timepoint])
par(mfrow = c(2, 2))
h(lcpm_lc475, col_list, "topleft")
```

```{r mds-lc476, fig.cap = "MDS plots of `LC476` log-CPM values over dimensions 1 and 2 and dimensioned 3 and 4 with samples coloured by treatment and timepoint. Distances on the plot correspond to the leading fold-change, which is the average (root-mean-square) log2-fold-change for the 500 genes most divergent between each pair of samples by default.", fig.asp = 1}
col_list <- list(
  "treatment" = treatment_colours[x_lc476$samples$treatment],
  "timepoint" = timepoint_colours[x_lc476$samples$timepoint])
par(mfrow = c(2, 2))
h(lcpm_lc476, col_list, "topleft")
```

## Summary

Datasets where samples do not cluster by experimental group may show little or no evidence of differential expression in the downstream analysis.
The above analysis suggests we may observe differential expression between timepoints, once the batch effect is properly accounted for, but not much between treatments.

# Differential expression analysis

This experiment has two levels of variability.
First, there is the variation from person to person, which we call the between-subject strata.
Then there is the variability of repeat measurements made on the same subject, the within-subject strata.
The between-subject variation is always expected to be larger than within-subject, because the latter is adjusted for baseline differences between the subjects.
Here the comparison between timepoints can be made within subjects, and hence should be more precise than the comparison between arms, which must be made between subjects.

<aside>

</aside>

## Creating a design matrix and contrasts

In this study, the primary interest is to see which genes are expressed at different levels between the different treatments and timepoints.
In our analysis, linear models are fitted to the data with the assumption that the underlying data is normally distributed.
To get started, a design matrix is set up with the `group` (i.e. the combination of `timepoint` and `treatment`).
There is also the unwanted variation between plates, which we account for by including a term for `plate_number` in the design matrix when [Using both plates] (this term is not required when just [Using LC476]).

<aside>
Of secondary interest is to see which genes have their expression associated with additional covariates (e.g., weight for age at endline). We will use different design matrices and contrasts to answer these questions.
</aside>

```{r}
design <- model.matrix(~0 + group + plate_number, x$samples)
colnames(design) <- sub("group|plate\\_number", "", colnames(design))

design_lc476 <- model.matrix(~0 + group, x_lc476$samples)
colnames(design_lc476) <- sub("group", "", colnames(design_lc476))
```

We wish to perform pairwise comparisons between `treatment`s, `timepoint`s, and their interactions, and we set these contrasts up using the `makeContrasts()` function from the `r BiocStyle::Biocpkg("limma")` package.

```{r}
contr_matrix <- makeContrasts(
  A.endline_vs_A.baseline = A.endline - A.baseline,
  B.endline_vs_B.baseline = B.endline - B.baseline,
  C.endline_vs_C.baseline = C.endline - C.baseline,
  A.endline_vs_B.endline = A.endline - B.endline,
  A.endline_vs_C.endline = A.endline - C.endline,
  B.endline_vs_C.endline = B.endline - C.endline,
  A_delta_vs_B_delta = (A.endline - A.baseline) - (B.endline - B.baseline),
  A_delta_vs_C_delta = (A.endline - A.baseline) - (C.endline - C.baseline),
  B_delta_vs_C_delta = (B.endline - B.baseline) - (C.endline - C.baseline),
  levels = colnames(design))

contr_matrix_lc476 <- makeContrasts(
  A.endline_vs_A.baseline = A.endline - A.baseline,
  B.endline_vs_B.baseline = B.endline - B.baseline,
  C.endline_vs_C.baseline = C.endline - C.baseline,
  A.endline_vs_B.endline = A.endline - B.endline,
  A.endline_vs_C.endline = A.endline - C.endline,
  B.endline_vs_C.endline = B.endline - C.endline,
  A_delta_vs_B_delta = (A.endline - A.baseline) - (B.endline - B.baseline),
  A_delta_vs_C_delta = (A.endline - A.baseline) - (C.endline - C.baseline),
  B_delta_vs_C_delta = (B.endline - B.baseline) - (C.endline - C.baseline),
  levels = colnames(design_lc476))
```

## Removing heteroscedascity from count data

It has been shown that for RNA-seq count data, the variance is not independent of the mean [@law2014voom] – this is true of raw counts or when transformed to log-CPM values.
Methods that model counts using a Negative Binomial distribution assume a quadratic mean-variance relationship.

In `r BiocStyle::Biocpkg("limma")`, linear modelling is carried out on the log-CPM values which are assumed to be normally distributed and the mean-variance relationship is accommodated using precision weights calculated by the `voom()` function.
`voom()` converts raw counts to log-CPM values by automatically extracting library sizes and normalisation factors.

<aside>
We actually use the more general `voomLmFit()` function from the `r BiocStyle::Biocpkg("edgeR")` package rather than `voom()` itself. `voomLmFit()` adapts the limma voom method [@law2014voom] to allow for loss of residual degrees of freedom due to exact zero counts [@lun2017no]. The loss residual df occurs when all the counts in a group are zero or when there are blocking factors that can fit zero counts exactly.
</aside>

The mean-variance relationship of log-CPM values for this dataset is shown in the left panels of Figure \@ref(fig:voom).
Typically, the voom-plot shows a decreasing trend between the means and variances resulting from a combination of technical variation in the sequencing experiment and biological variation amongst the replicate samples from different cell populations.
Experiments with high biological variation usually result in flatter trends, where variance values plateau at high expression values.
Experiments with low biological variation tend to result in sharp decreasing trends.

```{r voom, fig.cap = "Means (x-axis) and variances (y-axis) of each gene are plotted to show the dependence between the two before voom is applied to the data (left panels) and how the trend is removed after voom precision weights are applied to the data (right panels). The top row is when [Using both plates] and the bottom row is when just [Using LC476]. The plot on the left is created within the `voomLmFit()` function which extracts residual variances from fitting linear models to log-CPM transformed data. Variances are then rescaled to quarter-root variances (or square-root of standard deviations) and plotted against the average log2 count for each gene. The plot on the right is created using `plotSA()` which plots log2 residual standard deviations against mean log-CPM values. In both plots, each black dot represents a gene. On the left plot, the red curvse show the estimated mean-variance trend used to compute the voom weights (`voomLmFit()` makes a first estimate and refines it for the final estimate). On the right plot, the average log2 residual standard deviation estimated by the empirical Bayes algorithm is marked by a horizontal blue line and outlier variances are highlighted in red.", fig.asp = 1}
par(mfrow = c(2, 2))
vfit <- voomLmFit(x, design, x$samples$id, plot = TRUE)
vfit <- contrasts.fit(vfit, contrasts = contr_matrix)
efit <- eBayes(vfit, robust = TRUE)
plotSA(efit, main = "Both plates: Mean-variance trend")

vfit_lc476 <- voomLmFit(x_lc476, design_lc476, x_lc476$samples$id, plot = TRUE)
vfit_lc476 <- contrasts.fit(vfit_lc476, contrasts = contr_matrix_lc476)
efit_lc476 <- eBayes(vfit_lc476)
plotSA(efit_lc476, main = "LC476: Mean-variance trend")
```

## Fitting linear models for comparisons of interest

Linear modelling in `r BiocStyle::Biocpkg("limma")` fits a separate model to the expression values for each gene.
Next, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability [@smyth2004linear].
The model's residual variances are plotted against average expression values in Figure \@ref(fig:voom).

## Further normaliztion

Figure \@ref(fig:voom) shows that the variance of the expression levels is still somewhat dependent on the mean expression level, which is undesirable.
This suggests that the TMM normalization is insufficient to adjust for technical differences between samples, which means that subsequent analysis may be biased (e.g., 'differentially expressed genes' may be false positives).
This dependency is also observed in the MD plots^[An MD-plot, also known as an MA-plot, is a plot of log fold changes (differences) versus average log values (means).] shown in Figures \@ref(fig:md-plot) and \@ref(fig:md-plot-lc476).

```{r md-plot, fig.cap = "MD-plot of model coefficient when using both plates.", fig.asp = 1, results = "hide"}
par(mfrow = c(3, 3))
lapply(seq_len(ncol(efit)), function(j) {
  plotMD(
    efit,
    coef = j,
    xlim = range(efit$Amean),
    ylim = range(coef(efit)),
    cex = 0.5)
})
```

```{r md-plot-lc476, fig.cap = "MD-plot of model coefficient when just using LC476.", fig.asp = 1, results = "hide"}
par(mfrow = c(3, 3))
lapply(seq_len(ncol(efit_lc476)), function(j) {
  plotMD(
    efit_lc476,
    coef = j,
    xlim = range(efit_lc476$Amean),
    ylim = range(coef(efit_lc476)),
    cex = 0.5)
})
```

To correct this dependency we require additional, stronger normalization of the data.
I have opted for quantile normalization, which forces the entire empirical distribution of each sample to be identical[@bolstad2003comparison].
This approach was suggested to me by Matt Ritchie, Charity Law, and Gordon Smyth^[Email from Gordon (2020-09-20): "The uptick at high expression levels [in the MA plot but also true for the SA plot] does look a bit suspicious. I suggest switching to quantile normalization. It is heavier than TMM but still performs fine and may very well solve the issue. Wei's lab has used quantile normalization for lots of RNA-seq projects without any apparent ill effects."], authors of the the `r BiocStyle::Biocpkg("limma")` and `r BiocStyle::Biocpkg("edgeR")` packages.

Figure \@ref(fig:voom-with-quantile-normalization) shows that by quantile normalizing the data we successfully remove the dependency of the variance on the mean expression level.
Likewise, the removal of this dependency is also observed in the MD plots shown in Figures \@ref(fig:md-plot-quantile-normalized) and \@ref(fig:md-plot-lc476-quantile-normalized).

```{r voom-with-quantile-normalization, fig.cap = "Means (x-axis) and variances (y-axis) of each gene are plotted to show the dependence between the two before voom is applied to the data (left panels) and how the trend is removed after voom precision weights are applied to the data and the data are quantile normalized (right panels). The top row is when [Using both plates] and the bottom row is when just [Using LC476]. The plot on the left is created within the `voomLmFit()` function which extracts residual variances from fitting linear models to log-CPM transformed data. Variances are then rescaled to quarter-root variances (or square-root of standard deviations) and plotted against the average log2 count for each gene. The plot on the right is created using `plotSA()` which plots log2 residual standard deviations against mean log-CPM values. In both plots, each black dot represents a gene. On the left plot, the red curvse show the estimated mean-variance trend used to compute the voom weights (`voomLmFit()` makes a first estimate and refines it for the final estimate). On the right plot, the average log2 residual standard deviation estimated by the empirical Bayes algorithm is marked by a horizontal blue line and outlier variances are highlighted in red.", fig.asp = 1}
par(mfrow = c(2, 2))
vfit <- voomLmFit(
  x,
  design,
  x$samples$id,
  plot = TRUE,
  normalize.method = "quantile")
vfit <- contrasts.fit(vfit, contrasts = contr_matrix)
efit <- eBayes(vfit, robust = TRUE)
plotSA(efit, main = "Both plates: Mean-variance trend")

vfit_lc476 <- voomLmFit(
  x_lc476,
  design_lc476,
  x_lc476$samples$id,
  plot = TRUE,
  normalize.method = "quantile")
vfit_lc476 <- contrasts.fit(vfit_lc476, contrasts = contr_matrix_lc476)
efit_lc476 <- eBayes(vfit_lc476, robust = TRUE)
plotSA(efit_lc476, main = "LC476: Mean-variance trend")
```

```{r md-plot-quantile-normalized, fig.cap = "MD-plot of model coefficient when using both plates and the data are quantile normalized.", fig.asp = 1, results = "hide"}
par(mfrow = c(3, 3))
lapply(seq_len(ncol(efit)), function(j) {
  plotMD(
    efit,
    coef = j,
    xlim = range(efit$Amean),
    ylim = range(coef(efit)),
    cex = 0.5)
})
```

```{r md-plot-lc476-quantile-normalized, fig.cap = "MD-plot of model coefficient when just using LC476  and the data are quantile normalized.", fig.asp = 1, results = "hide"}
par(mfrow = c(3, 3))
lapply(seq_len(ncol(efit_lc476)), function(j) {
  plotMD(
    efit_lc476,
    coef = j,
    xlim = range(efit_lc476$Amean),
    ylim = range(coef(efit_lc476)),
    cex = 0.5)
})
```

# Introduction to results

I provide results [Using both plates] and just [Using LC476].
The two analyses use the same methodology but [Using both plates] makes full use of all the data generated.
I recommend the results from [Using both plates], but leave the final choice to you.

# Using both plates

Please see [`output/DEGs/both_plates/`](../output/DEGs/both_plates/) for spreadsheets of these DEG lists, including results of GO and KEGG analyses using the `goana()` and `kegga()` functions from the `r BiocStyle::Biocpkg("limma")` package, as well as heatmaps and Venn diagrams of the DEG lists and interactive `r BiocStyle::Biocpkg("Glimma")` plots of the differential expression results.
The heatmaps and right panels of the `r BiocStyle::Biocpkg("Glimma")` plots use the batch-corrected log-CPMs.
Group labels may be abbreviated in the `r BiocStyle::Biocpkg("Glimma")` plots.
For example, when looking at the effect of `Stunted24`, the group `5.A.F` denotes samples on plate `LC475` (**5**) with treatment `A` (**A**) that were not stunted (**F**ALSE); group `6.B.T` denotes samples on `LC476` (**6**) with treatment `B` (**B**) that were stunted (**T**RUE); etc.

```{r}
deg_dir <- here("output", "DEGs", "both_plates")
dir.create(deg_dir, recursive = TRUE)
```

## Figure 1

### Endline pairwise analyses between treatments

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
efit_ <- efit[, c(
    "A.endline_vs_B.endline",
    "A.endline_vs_C.endline",
    "B.endline_vs_C.endline")]

dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
x_ <- x[, x$sample$timepoint == "endline"]
o <- order(
  x_$samples$plate_number,
  x_$samples$treatment,
  x_$samples$sex,
  x_$samples$id)
x_ <- x_[, o]
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.endline_treatment_comparisons",
  suffix = NULL,
  groups = "treatment",
  sample.cols = "treatment_colours",
  annotation_col = c("plate_number", "treatment", "sex"))
```

#### Adjusted by sex

Tom and Santa also requested the analysis of endline pairwise analyses between treatments be adjusted by sex.
This is done by including `sex` in the design matrix. 

```{r}
design_with_sex <- model.matrix(~0 + group + sex, x$samples)
colnames(design_with_sex) <- sub("sex|group", "", colnames(design_with_sex))
contr_matrix_with_sex <- makeContrasts(
  A.endline_vs_A.baseline = A.endline - A.baseline,
  B.endline_vs_B.baseline = B.endline - B.baseline,
  C.endline_vs_C.baseline = C.endline - C.baseline,
  A.endline_vs_B.endline = A.endline - B.endline,
  A.endline_vs_C.endline = A.endline - C.endline,
  B.endline_vs_C.endline = B.endline - C.endline,
  A_delta_vs_B_delta = (A.endline - A.baseline) - (B.endline - B.baseline),
  A_delta_vs_C_delta = (A.endline - A.baseline) - (C.endline - C.baseline),
  B_delta_vs_C_delta = (B.endline - B.baseline) - (C.endline - C.baseline),
  levels = colnames(design_with_sex))
vfit_with_sex <- voomLmFit(
  x,
  design_with_sex,
  x$samples$id,
  plot = FALSE, 
  normalize.method = "quantile")
vfit_with_sex <- contrasts.fit(vfit_with_sex, contrasts = contr_matrix_with_sex)
efit_with_sex <- eBayes(vfit_with_sex, robust = TRUE)
```

The results are summarised below.

```{r}
efit_ <- efit_with_sex[, c(
    "A.endline_vs_B.endline",
    "A.endline_vs_C.endline",
    "B.endline_vs_C.endline")]

dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r}
# Create outputs
x_ <- x[, x$sample$timepoint == "endline"]
o <- order(
  x_$samples$plate_number,
  x_$samples$treatment,
  x_$samples$sex,
  x_$samples$id)
x_ <- x_[, o]
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.endline_treatment_comparisons",
  suffix = "adjusted_by_sex",
  groups = "treatment",
  sample.cols = "treatment_colours",
  annotation_col = c("plate_number", "treatment", "sex"))
```

## Figure 2

### Endline vs. baseline

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
efit_ <- efit[, c(
    "A.endline_vs_A.baseline",
    "B.endline_vs_B.baseline",
    "C.endline_vs_C.baseline")]
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
x_ <- x
o <- order(
  x_$samples$plate_number,
  x_$sample$treatment,
  x_$sample$timepoint,
  x_$sample$sex,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$treatment.timepoint <- paste(
  x_$sample$treatment,
  x_$samples$timepoint,
  sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.timepoint_comparisons",
  suffix = NULL,
  groups = "treatment.timepoint",
  sample.cols = "treatment_colours",
  annotation_col = c("plate_number", "treatment", "timepoint", "sex"))
```

#### Check for differences in age at enrolment

```{r}
tmp <- unique(x$samples[, c("id", "treatment", "age0")])
tmp$age0_cut <- cut(tmp$age0, breaks = c(8.5, 10, 11.5))
```

Figure \@ref(fig:age0-density) shows the distribution of age at enrolment (`age0`) by treatment arm.
There is no significant difference in `age0` by treatment arm (ANOVA, $P =$ `r number(summary(aov(tmp$age0 ~ tmp$treatment))[[1]][["Pr(>F)"]][1], accuracy = 0.001)`) nor is there an association when `age0` is binned as $(8.5-10]$ or $(10-11.5]$ ($\chi^2$-test $P =$ `r number(chisq.test(tmp$treatment, tmp$age0_cut)$p.value, accuracy = 0.01)`).

```{r age0-density, fig.cap = "Density estimates of `age0` in each treatment arm. The density curves were scaled such that the area under each curve corresponds to the total number of subjects in each treatment arm."}
ggplot(tmp, aes(x = age0, y = ..count.., fill = treatment)) + 
  geom_density(alpha = 1 / 2) +
  scale_fill_manual(values = treatment_colours) +
  theme_minimal_hgrid() +
  ylab("scaled density") +
  guides(fill = guide_legend(override.aes = list(linetype = 0))) +
  theme(axis.line.x = element_blank()) + 
  geom_vline(xintercept = 10, lty = 2)
```

## Figure 3

### Gene expression by Height for Age at endline

```{r}
tmp <- unique(x$samples[, c("id", "treatment", "HAZ24", "Stunted24")])
```

Figure \@ref(fig:HAZ24-density) shows the distribution of height for age at endline (`HAZ24`) by treatment arm.
There is `r sum(is.na(tmp$HAZ24))` samples without a `HAZ24` measurement (id: `r glue::glue_collapse(tmp$id[is.na(tmp$HAZ24)], sep = ", ", last = ", and ")`).

```{r HAZ24-density, fig.cap = "Density estimates of `HAZ24` in each treatment arm. The density curves were scaled such that the area under each curve corresponds to the total number of subjects in each treatment arm. Samples to the left of the vertical dashed line are `Stunted24`."}
ggplot(tmp, aes(x = HAZ24, y = ..count.., fill = treatment)) + 
  geom_density(alpha = 1 / 2) +
  scale_fill_manual(values = treatment_colours) +
  theme_minimal_hgrid() +
  ylab("scaled density") +
  guides(fill = guide_legend(override.aes = list(linetype = 0))) +
  theme(axis.line.x = element_blank()) + 
  geom_vline(xintercept = -2, lty = 2)
```

Stunting at endline (`Stunted24`) is defined as `HAZ24` $< -2$.
The table below classifies subjects by `treatment` and `Stunted24`.

```{r}
knitr::kable(
  janitor::tabyl(tmp, Stunted24, treatment) %>%
    janitor::adorn_title("combined"))
```

#### `HAZ24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `HAZ24`

Contrasts for association of gene expression with `HAZ24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$HAZ24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + HAZ24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(HAZ24 = HAZ24, levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$HAZ24,
  x_$sample$Stunted24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
    sub("LC47", "", x_$sample$plate_number),
    x_$sample$treatment,
    ifelse(x_$sample$Stunted24, "T", "F"),
    sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.HAZ24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Stunted24",
  annotation_col = c("plate_number", "treatment", "HAZ24", "Stunted24"))
```

#### `Stunted24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `Stunted24`.

Contrasts for association of gene expression with `Stunted24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$Stunted24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + Stunted24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(
  Stunted24 = Stunted24TRUE,
  levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
    x_$sample$plate_number,
    x_$sample$treatment,
    x_$sample$HAZ24,
    x_$sample$Stunted24,
    x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
    sub("LC47", "", x_$sample$plate_number),
    x_$sample$treatment,
    ifelse(x_$sample$Stunted24, "T", "F"),
    sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.Stunted24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Stunted24",
  annotation_col = c("plate_number", "treatment", "HAZ24", "Stunted24"))
```

### Gene expression by Weight for Height at endline

```{r}
tmp <- unique(x$samples[, c("id", "treatment", "WHZ24", "Wasted24")])
```

Figure \@ref(fig:WHZ24-density) shows the distribution of weight for height at endline (`WHZ24`) by treatment arm.
There is `r sum(is.na(tmp$WHZ24))` samples without a `WHZ24` measurement (id: `r glue::glue_collapse(tmp$id[is.na(tmp$WHZ24)], sep = ", ", last = ", and ")`).

```{r WHZ24-density, fig.cap = "Density estimates of `WHZ24` in each treatment arm. The density curves were scaled such that the area under each curve corresponds to the total number of subjects in each treatment arm. Samples to the left of the vertical dashed line are `Wasted24`."}
ggplot(tmp, aes(x = WHZ24, y = ..count.., fill = treatment)) + 
  geom_density(alpha = 1 / 2) +
  scale_fill_manual(values = treatment_colours) +
  theme_minimal_hgrid() +
  ylab("scaled density") +
  guides(fill = guide_legend(override.aes = list(linetype = 0))) +
  theme(axis.line.x = element_blank()) +
  geom_vline(xintercept = -2, lty = 2)
```

Wasting at endline (`Wasted24`) is defined as `WHZ24` $< -2$.
The table below classifies subjects by `treatment` and `Wasted24`.

```{r}
knitr::kable(
  janitor::tabyl(tmp, Wasted24, treatment) %>%
    janitor::adorn_title("combined"))
```

#### `WHZ24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `WHZ24`.

Contrasts for association of gene expression with `WHZ24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$WHZ24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + WHZ24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(WHZ24 = WHZ24, levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$WHZ24,
  x_$sample$Wasted24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
    sub("LC47", "", x_$sample$plate_number),
    x_$sample$treatment,
    ifelse(x_$sample$Wasted24, "T", "F"),
    sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.WHZ24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Wasted24",
  annotation_col = c("plate_number", "treatment", "WHZ24", "Wasted24"))
```

#### `Wasted24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `Wasted24`.

Contrasts for association of gene expression with `Wasted24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$Wasted24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + Wasted24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(
  Wasted24 = Wasted24TRUE,
  levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$WHZ24,
  x_$sample$Wasted24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
  sub("LC47", "", x_$sample$plate_number),
  x_$sample$treatment,
  ifelse(x_$sample$Wasted24, "T", "F"),
  sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.Wasted24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Wasted24",
  annotation_col = c("plate_number", "treatment", "WHZ24", "Wasted24"))
```

### Gene expression by Weight for Age at endline

```{r}
tmp <- unique(x$samples[, c("id", "treatment", "WAZ24", "Underweight24")])
```

Figure \@ref(fig:WAZ24-density) shows the distribution of weight for age at endline (`WAZ24`) by treatment arm.
There is `r sum(is.na(tmp$WAZ24))` samples without a `WAZ24` measurement (id: `r glue::glue_collapse(tmp$id[is.na(tmp$WAZ24)], sep = ", ", last = ", and ")`).

```{r WAZ24-density, fig.cap = "Density estimates of `WAZ24` in each treatment arm. The density curves were scaled such that the area under each curve corresponds to the total number of subjects in each treatment arm. Samples to the left of the vertical dashed line are `Underweight24`."}
ggplot(tmp, aes(x = WAZ24, y = ..count.., fill = treatment)) + 
  geom_density(alpha = 1 / 2) +
  scale_fill_manual(values = treatment_colours) +
  theme_minimal_hgrid() +
  ylab("scaled density") +
  guides(fill = guide_legend(override.aes = list(linetype = 0))) +
  theme(axis.line.x = element_blank()) +
  geom_vline(xintercept = -2, lty = 2)
```

Underweight at endline (`Underweight24`) is defined as `WAZ24` $< -2$.
The table below classifies subjects by `treatment` and `Underweight24`.

```{r}
knitr::kable(
  janitor::tabyl(tmp, Underweight24, treatment) %>%
    janitor::adorn_title("combined"))
```

#### `WAZ24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `WAZ24`

Contrasts for association of gene expression with `WAZ24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$WAZ24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + WAZ24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(WAZ24 = WAZ24, levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$WAZ24,
  x_$sample$Underweight24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
  sub("LC47", "", x_$sample$plate_number),
  x_$sample$treatment,
  ifelse(x_$sample$Underweight24, "T", "F"),
  sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.WAZ24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Underweight24",
  annotation_col = c("plate_number", "treatment", "WAZ24", "Underweight24"))
```

#### `Underweight24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `Underweight24`.

Contrasts for association of gene expression with `Underweight24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$Underweight24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + Underweight24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(
  Underweight24 = Underweight24TRUE,
  levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$WAZ24,
  x_$sample$Underweight24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
  sub("LC47", "", x_$sample$plate_number),
  x_$sample$treatment,
  ifelse(x_$sample$Underweight24, "T", "F"),
  sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.Underweight24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Underweight24",
  annotation_col = c("plate_number", "treatment", "WAZ24", "Underweight24"))
```

## Figure 4

### Gene expression by Haemoglobin at endline

```{r}
tmp <- unique(x$samples[, c("id", "treatment", "hb24", "Anaemic24")])
```

Figure \@ref(fig:hb24-density) shows the distribution of weight for age at endline (`hb24`) by treatment arm.
There is `r sum(is.na(tmp$hb24))` samples without a `hb24` measurement (id: `r glue::glue_collapse(tmp$id[is.na(tmp$hb24)], sep = ", ", last = " and ")`).

```{r hb24-density, fig.cap = "Density estimates of `hb24` in each treatment arm. The density curves were scaled such that the area under each curve corresponds to the total number of subjects in each treatment arm. Samples to the left of the vertical dashed line are `Anaemic24`."}
ggplot(tmp, aes(x = hb24, y = ..count.., fill = treatment)) + 
  geom_density(alpha = 1 / 2) +
  scale_fill_manual(values = treatment_colours) +
  theme_minimal_hgrid() +
  ylab("scaled density") +
  guides(fill = guide_legend(override.aes = list(linetype = 0))) +
  theme(axis.line.x = element_blank()) + 
  geom_vline(xintercept = 11, lty = 2)
```

Anaemic at endline (`Anaemic24`) is defined as `hb24` $< -2$.
The table below classifies subjects by `treatment` and `Anaemic24`.

```{r}
knitr::kable(
  janitor::tabyl(tmp, Anaemic24, treatment) %>%
    janitor::adorn_title("combined"))
```

#### `hb24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `hb24`

Contrasts for association of gene expression with `hb24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$hb24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + hb24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(hb24 = hb24, levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$hb24,
  x_$sample$Anaemic24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
  sub("LC47", "", x_$sample$plate_number),
  x_$sample$treatment,
  ifelse(x_$sample$Anaemic24, "T", "F"),
  sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.hb24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Anaemic24",
  annotation_col = c("plate_number", "treatment", "hb24", "Anaemic24"))
```

#### `Anaemic24`

We subset the data to just the endline `timepoint` and update the original design matrix to add `Anaemic24`.

Contrasts for association of gene expression with `Anaemic24`, accounting for `group` and `plate_number`, are set up in `r BiocStyle::Biocpkg("limma")` using the `makeContrasts()` function.

```{r, results = "hide"}
x_ <- x[, !is.na(x$samples$Anaemic24) & x$samples$timepoint == "endline"]

design_ <- model.matrix(~0 + group + Anaemic24 + plate_number, x_$samples)
colnames(design_) <- sub("plate\\_number|group", "", colnames(design_))

contr_matrix_ <- makeContrasts(
  Anaemic24 = Anaemic24TRUE,
  levels = colnames(design_))

vfit_ <- voomLmFit(
  x_,
  design_,
  x_$samples$id,
  plot = FALSE,
  normalize.method = "quantile")
vfit_ <- contrasts.fit(vfit_, contrasts = contr_matrix_)
efit_ <- eBayes(vfit_, robust = TRUE)
```

For a quick look at differential expression levels, the number of significantly up- and down-regulated genes can be summarised in a table.
Significance is defined using an adjusted p-value cutoff that is set at 5% by default.

```{r}
dt <- decideTests(efit_)
knitr::kable(
  t(summary(dt)),
  caption = "Number of DEGs per comparison. Each row corresponds to a comparison and each column corresponds to the number of downregulated genes (`Down`), the number of non-differentially expressed genes (`NotSig`), the number of upregulated genes (`Up`)")
```

```{r, message = FALSE}
# Create outputs
o <- order(
  x_$sample$plate_number,
  x_$sample$treatment,
  x_$sample$hb24,
  x_$sample$Anaemic24,
  x_$sample$id)
x_ <- x_[, o]
x_$samples$groups <- paste(
  sub("LC47", "", x_$sample$plate_number),
  x_$sample$treatment,
  ifelse(x_$sample$Anaemic24, "T", "F"),
  sep = ".")
createDEGOutputs(
  outdir = deg_dir,
  efit = efit_,
  x = x_,
  prefix = "both_plates.Anaemic24",
  suffix = NULL,
  groups = "groups",
  sample.cols = "Anaemic24",
  annotation_col = c("plate_number", "treatment", "hb24", "Anaemic24"))
```

```{r}
knitr::knit_exit()
```

# Using LC476

**TODO**: Re-write/remove and replace with analysis for Figures 1 - 4

It was decided to only use data from plate `LC476` for the differential expression analysis.
The reasoning being that we then don't have to deal with the batch effect associated with the `plate_number`.
Doing so slightly simplifies the analysis (as we don't have to block on `plate_number`) but may or may not also reduce our power to detect differential expression.
Briefly, for a given comparison a gene may be:

1. Not DE in either batch. These genes will not be detected as DE in an analysis of either batch or both batches (excepting a small number of false positives).
2. DE in both batches. These genes will be detected as DE in an analysis of either batch or both batches.
3. DE in one of the batches but not the other. These genes will be detected as DE in a analysis of the relevant batch but are likely not be detected as DE in an analysis of both batches because the signal is inconsistent between batches.
4. Weakly but consistently DE in each batch, i.e. not reaching statistical significance in either batch but with a consistent different between batches. These genes *may* be detected as DE in an analysis of both batches.

By only analysing `LC476` is that we lose our ability to detect genes in group 4.
The number of DE genes will therefore likely be lower than when analysing both batches, but with arguably a simpler interpretation of the results or slighly increased confidence in the results due to the avoidance of the batch effect.

```{r}
x <- x_[, x_$samples$plate_number == "LC476"]
```

## Removing genes that are lowly expressed

We repeat the process of filtering out lowly expressed genes with the `filterByExpr()` function in the `r BiocStyle::Biocpkg("edgeR")` package.

```{r}
keep.exprs <- filterByExpr(x, group = x$samples$group)
x_ <- x
x <- x[keep.exprs, , keep.lib.sizes = FALSE]
```

This retains `r scales::number(sum(keep.exprs), big.mark = ",")` genes for DE analysis.

# Additional information {.appendix}

The following are available on request:

- Full CSV tables of any data presented.
- PDF/PNG files of any static plots.

# Session info {.appendix}

<summary>The analysis and this document were prepared using the following software (click triangle to expand)</summary>
<details>

```{r}
sessioninfo::session_info()
```

</details>

# TODOs

- [ ] Interactions (delta coefficients)
